Do investors in clean energy ETFs herd? The role of climate risks

### 2025-06-25 alt_energy_herding.qmd
added from comments

Exchange-traded funds (ETFs) are a key feature of green financial products. They are a collection of investments that often tracks an underlying performance of an asset or index. In the universe of environment, social, and governance (ESG) investments, clean ETFs serve as a tool for environmentally conscious investors to identify and invest in environmentally friendly companies [@briere2023] engaged in the transition towards a cleaner production and a low-carbon economy. Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing one in 2022 [@decclesia2024]. This is not surprising given that the clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products have escalated enormously [@naqvi2022]. That is, the growth of green assets under management is likely to continue. Furthermore, the limited availability of data on ESG complying investment tools [@avramov2022; @nguyen2025] justifies the use of energy ETFs as best candidates of green assets. However, given that the inclusion of climate sustainability in investment decisions is a recent phenomenon, it is not clear what the actual impact will be in the long-run.

@@

Following the relevant literature [@christie1995; and @chang2000], we compute dispersion of the $i^{th}$ ETF from the market return, which is known as the Cross Sectional Absolute Deviation  ($CSAD_t$) measure. Empirically the $CSAD_t$ is defined in the following manner:

@@

>>>>

Energy market has been in the epicentre of retail investors and portfolio managers at least over the last two decades. Official estimates claim that global energy investment is expected to surpass USD 3 trillion for the first time in 2024, with USD 2 trillion going to clean energy technologies and infrastructure [@iea2024]. There is a general consensus in the literature that the most important event that has shaped the current landscape in the energy sector is the emergence of the energy portfolio with the rise of green energy as an alternative energy resource and an asset class. Price volatility of fossil fuels on the one hand and decarbonisation-related goals dictated by international agreements on the other hand are heavily responsible for the growing interest in green energy market. Responding to the increasing awareness about the environment and the need for more green capital, financial markets have also attracted a large portion of retail and institutional investors’ fund flows [@rizvi2022green; @naqvi2021there].

Needless to mention that the rise of energy popularity as an alternative investment class has also been fueled by the fictionalization of commodities markets. Commodities have become an effective investment tool for investment community as it allows investors to participate in the market developments across various physical products such as agricultural products, silver, gold, natural gas, oil etc and more recently clean energy products without being obliged to own the product directly [@gagnon2020they]. Several academic studies have highlighted substantial diversification benefits of commodities when added to a portfolio of stocks and bonds [see inter alia, @naqvi2022].

Green energy sectors like solar, wind, geothermal, biofuels, biomass, wave, hydro and tidal energies are considered by investors, authorities and policy-makers as the substitutes for fossil-fuels energy that increase CO2 emissions [@shahzad2020clean]. Moreover, as @naqvi2022 point out that "the alternative energy sector is expected to grow faster over the next decades due to the depletion of fossil fuel reserves and the technological advance that has made green energy feasible".

Although green energy investment has witnessed an exponential growth over the years, there is a lot of improvement that can be done in terms of investors’ participation, both retail and institutional. To achieve the goal of larger capital commitment to green energy projects it is important to gain investors’ confidence in financial products that are linked to green energy such as Exchange-traded funds(ETFs).

ETFs are a key feature of green financial products. They are a collection of investments that often tracks an underlying performance of an asset or index. In the universe of environment, social, and governance (ESG) investments, clean ETFs serve as a tool for environmentally conscious investors to identify and invest in environmentally friendly companies [@briere2023] engaged in the transition towards a cleaner production and a low-carbon economy. Among the ESG ETFs, the Clean Energy (CE) ETFs have been the best-performing one in 2022 [@decclesia2024]. This is not surprising given that the clean energy transition represents one of the largest multi-decade secular growth opportunities.  After the inclusion of Green energy financing in the list of United Nations Sustainability Goals (SDGs) as SDG 7, the role, importance, and visibility of green financial products have escalated enormously [@naqvi2022]. That is, the growth of green assets under management is likely to continue. Furthermore, the limited availability of data on ESG complying investment tools [@avramov2022; @nguyen2025] justifies the use of energy ETFs as best candidates of green assets. However, given that the inclusion of climate sustainability in investment decisions is a recent phenomenon, it is not clear what the actual impact will be in the long-run.

@@

More broadly, studies that focus on herding can be broadly classified into two main categories:  on the one hand there are studies that examine institutional investor herding and analyst herding using transaction-level data [@lakonishok1992; and @sias2004institutional, among others]. On the other hand, there are studies that explore herding towards the market consensus employing individual returns and market returns [@christie1995; and @chang2000]. In the latter case, the most widely employed method is the cross-sectional standard deviation of returns proposed by @christie1995. The intuition behind this approach is that herding toward the market is signaled when return dispersion is relatively low. Rational asset pricing theory predicts that when stocks respond to market movements in various degrees then dispersion will be linearly related to market returns [see also, @hwang2004market]. However, when investors tend to mimic each other and herd towards the market the linear relation between returns dispersion and aggregate market returns is no longer valid. Under these conditions, fitting a non-linear regression specification between a measure of returns dispersion and aggregate market return appears as a more viable choice. Therefore, our approach to measure herding activity is based on the non-linear regression model by @chang2000.

Following this relevant literature, we compute dispersion of the $i^{th}$ ETF from the market return, which is known as the Cross Sectional Absolute Deviation  ($CSAD_t$) measure. Empirically the $CSAD_t$ is defined in the following manner:

@@

Furthermore, higher transition risk means greater uncertainty for future pathway towards a cleaner economy. Under these conditions, green assets such as green stocks or ETFs are expected to perform better and can serve as hedging tools [see inter alia @pastor]. As a result, investors allocate their capital to available green assets thus resulting in higher returns’ dispersion with respect to the aggregate market. Stated differently, investors move their capital away from market consensus across green ETFs ruling out the possibility of herding behavior.
